---
title: "Cartography Assignment"
description: |
  Maps for life expectacy in Africa and South America
author:
  - name: Grace Kumaishi, Julia Parish
    url: https://github.com/juliaparish
    orchid_id: 0000-0002-8323-6574
date: 2021-10-15
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
    toc: true # add a table of contents
    toc_float: true #float it (positioned on the left)
categories:
  - R
  - GIS
preview: images/combined_lifeexp.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r, include = FALSE}
# Install required packages
library(sf)
library(spData)
library(RColorBrewer)
library(rnaturalearth)
library(rnaturalearthdata)
library(tidyverse)
library(tmap)    # for static and interactive maps
library(ggplot2) # tidyverse data visualization package
library(rgeos)
```

This cartography project was completed as an assignment for my Master's program course, Environmental Data Science 233: Spatial Analysis. It was collaboration with G. Kumaishi. 

## Project Goal

Create a map of life expectancy for South America and Africa.

### Step One
Create two separate maps for life expectancy data of South America and Africa using `spData` and `tmap`.

##### *Life Expectancy in Africa:*

```{r fig.cap="Life Expectancy in Africa", fig.alt = "Map of life expectancy for African countries. Life expectancy is visualized with five-year increments." , fig.align = "center"}
### Africa ###

world_africa <- world[world[["continent"]] == "Africa", ] # create a subset of the world dataset of just the African continent

lifeexp_africa <- tm_shape(world_africa) + # store map in order to combine
  tm_polygons(col = "lifeExp", # select life expectancy data
            palette = "YlOrBr", # Yellow Orange Brown color palette from RColorBrewer
            style = "pretty", # familiar GIS discrete options available such as quantile, equal, jenks, etc.
            title = "Life Expectancy (years)*", # legend title lives within tm_polygons
            textNA = "No Data") +   # converted "Missing" legend default to "No Data"
  tm_layout(inner.margins = c(0.1, 0.1, 0.1, 0.1), # margins between 0-1 (bottom, left, top, right)
            legend.title.size = 1,
            legend.text.size = .8,
            bg.color = "lightblue", # background color
            title = "Africa") + # map title lives under tm_layouts
  tm_compass(type = "4star", # other compass styles available
            show.labels = 1, # specifies compass label options
            size = 3,
            position = c("right", "top")) +
  tm_credits("Life expectancy data for the\nAfrican continent uses 5-year\nclassification breaks.\nSource: SpData \n \n*Note that the color scale differs\nbetween continents.", # credit data source including github link; \n creates text break
            size = .5, # can't figure out how to make size bigger...
            position = c("right", "bottom"), 
            align = "right", # only applicable if using multiple lines
            width = NA) # 1 is the whole map width, NA is determined by width of text

lifeexp_africa
```

##### *Life Expectancy in South America:*

```{r fig.cap="Life Expectancy in South America", fig.alt = "Map of life expectancy for South American countries. Life expectancy is visualized with two-year increments.", fig.align = "center"}
### South America ###

world_sa <- world[world[["continent"]] == "South America", ] # create a subset of the world dataset of just the South American continent

lifeexp_sa <- tm_shape(world_sa) + # store map in order to combine
  tm_polygons(col = "lifeExp", # select life expectancy data
            palette = "YlOrBr", # Yellow Orange Brown color palette from RColorBrewer
            n = 5, # specify number of breaks
            style = "pretty", # familiar GIS discrete options available such as quantile, equal, jenks, etc.
            title = "Life Expectancy (years)*", # legend title lives within tm_polygons
            textNA = "No Data") + # converted "Missing" legend default to "No Data"
  tm_layout(inner.margins = c(0.1, 0.25, 0.1, 0.25), # margins between 0-1 (bottom, left, top, right)
            legend.position = c("right", "bottom"), 
            legend.title.size = 1,
            legend.text.size = .8,
            bg.color = "lightblue", # background color
            title = "South America") + # map title lives under tm_layouts
  tm_compass(type = "4star", # other compass styles available
             show.labels = 1, # specifies compass label options
             size = 3,
             position = c("right", "top")) +
  tm_credits("Life expectancy data\nfor the South American continent uses\n2-year classification breaks.\nSource: SpData \n \n*Note that the color scale differs\nbetween continents.", # credit data source including github link; \n creates text break
             position = c("left", "bottom"), 
             align = "left",
             size = .5) + 
  tm_credits("", # explain varying color scales b/w two continents; \n creates text break
             position = c("right", "bottom"), 
             align = "right",
             size = .3)
  
lifeexp_sa
```

### Step Two
Combine the two maps using `tmap_arrange` to compare and contrast life expectancy rates.

##### *Life Expectancy in South America and Africa*:\*

```{r fig.cap="Fig 3: A Comparision of Life Expectancy between South America and Africa", fig.alt = "A figure comparing life expectancy between South America and Africa. South America is visualized with two-year increments, while Africa is visualized with 5-year increments.", fig.align = "center"}
combined_lifeexp <- tmap_arrange(lifeexp_sa,lifeexp_africa) # combine the two life expectancy maps using tmap_arrange

combined_lifeexp
```

### Conclusion

This project was to practice creating maps in R using `sf`, `tmap`, and `spData` packages. 

