---
title: "What is the effect of sea surface temperature on coral bleaching?"
description: |
  A statisical analysis of the impact of the 2011 Tōhoku tsunami on albatross populations in Hawaii.
author:
  - name: Julia Parish, Desik Somasundaram, Ryan Munnikhuis, and Cullen Molitor
    url: https://github.com/desik23/eds-213-group-project
date: 2021-11-13
bibliography: bibliography.bib
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
    toc: true # add a table of contents
    toc_float: true #float it (positioned on the left)
categories:
  - R
draft: true
preview: images/rowleycredit-scott-portelli.jpg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(metajam)
library(here)
library(tidyverse)
library(lubridate)
library(janitor)
library(vegan)
library(zoo)
```

# EDS 213 Metadata Group Project
Our group discovered the Scott Reef and Rowley Shoals Coral Bleaching Data dataset while searching the DataOne repository for “coral bleaching.” This particular dataset focuses on long-term monitoring data from 1994 to 2017 at reef slope habitats off the coast of northwestern Australia. The metadata includes the reef system, date range, taxonomy, habitat type, coral coverage, and dataset methods. 

```{r error =TRUE}
# assign data url to access coral data from DataOne then download
data_url <- "https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fedi%2F952%2F1%2Ff6212784c45d0a077f2c863868d22c4b"
# If data is akready up to date, this will throw an error
# error=T in header will move past this issue
download_d1_data(data_url, dir_name = "coral_data", path = ".")
```

```{r}
# assign data path
data_path <- here::here("coral_data")
```

```{r}
# Read in data with metajam
coral_list <- read_d1_files(data_path)
```

```{r}
# Pick out the data read in above and clean it up
corals <- coral_list$data %>% 
  janitor::clean_names() %>% 
  mutate(year_decimal = format(date_decimal(year_decimal), "%Y-%m-%d"),
         month = month(year_decimal),
         year = year(year_decimal)) %>% 
  rename(date = year_decimal)
```

```{r}
# Read in  ONI to be joined to coral data
oni <- read.table(
  "https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt",
  header = T) %>%
  dplyr::mutate(date = as.Date(ISOdate(YR, MON, 1)),
                date_start = as.Date(ISOdate(YR, MON, 1)),
                date_end = ceiling_date(date_start, "month")) %>%
  dplyr::rename(oni_anomaly = ANOM,
                month = MON,
                year = YR) %>% 
  dplyr::select(year, month, oni_anomaly, date_start, date_end) %>% 
  dplyr::mutate(roll_3_month_mean = zoo::rollmean(x = oni_anomaly, k = 3, fill = NA, align = "right")) %>% 
  filter(date_start > ymd("1994-09-01"))
```

```{r}
# Calculate the Simpson's diversity index 
# join corals and ONI data
coral_oni <- corals %>% 
  pivot_longer(cols = 5:18, names_to = "species", values_to = "cover") %>% 
  dplyr::group_by(system, reef, year, month, location) %>% 
  dplyr::summarise(simpson_index = vegan::diversity(cover, index = "simpson")) %>% 
  dplyr::left_join(oni %>% select(-date_start, -date_end)) %>% 
  dplyr::mutate(date = lubridate::make_date(year = year, month = month, day = 1))
```

```{r}
#Un-comment line below to write the updated csv file
readr::write_csv(coral_oni, here("coral_diversity_oceanic_nino_index.csv"))
```

```{r fig.height=4, fig.width=9}
# Plot the coral diversity data with ONI color bar
ggplot() +
  ggplot2::geom_rect(
    data = oni,
    aes(xmin = date_start, xmax = date_end, ymin = -Inf, ymax = .3, fill = oni_anomaly)) +
  ggplot2::scale_fill_viridis_c(
    option = "plasma",
    guide = guide_colorbar(direction = "horizontal", title.position = "top",
                           order = 2, barheight = unit(.2, "cm"))) +
  scale_x_date(date_breaks = "2 year", date_labels = "%Y", expand = expansion(mult = c(0,0))) +
  geom_line(data = coral_oni, size = 1,
              aes(x = date, y = simpson_index, color = location)) +
  scale_color_viridis_d() +
  ggplot2::guides(color = guide_legend(order = 1)) +
  labs(fill = "Oceanic Ni\u00f1o Index", color = "Sites", 
       x = "Date", y = "Simpsons Diversity Index") +
  theme_minimal()
```

```{r}
# In the future think about an appropriate model and explore the relationship
# mod <- lm(simpson_index ~ roll_3_month_mean + location, data = coral_oni)
# summary(mod)
```

## Plot
```{r fig.height=4, fig.width=9}
# Plot the coral diversity data with ONI color bar
ggplot2::ggplot() +
  ggplot2::geom_rect(
    data = oni,
    aes(xmin = date_start, xmax = date_end, ymin = -Inf, ymax = .3, fill = oni_anomaly)) +
  ggplot2::scale_fill_viridis_c(
    option = "plasma",
    guide = guide_colorbar(direction = "horizontal", title.position = "top",
                           order = 2, barheight = unit(.2, "cm"))) +
  ggplot2::scale_x_date(date_breaks = "2 year", date_labels = "%Y", expand = expansion(mult = c(0,0))) +
  ggplot2::geom_line(data = coral_oni, size = 1,
                     aes(x = date, y = simpson_index, color = location)) +
  ggplot2::scale_color_viridis_d() +
  ggplot2::guides(color = guide_legend(order = 1)) +
  ggplot2::labs(fill = "Oceanic Ni\u00f1o Index", color = "Sites", 
                x = "Date", y = "Simpsons Diversity Index") +
  ggplot2::theme_minimal()
```

## Data Sources and Download

**1.** James P Gilmour, Kylie L Cook, and Nicole M Ryan. 2021. Scott Reef and Rowley Shoals Coral Bleaching Data. Environmental Data Initiative.<br>
[LTER Data Access](https://pasta.lternet.edu/package/metadata/eml/edi/952/1.)<br>
- **Download Method:** metajam data_downloade API
- **Spatial Coverage**: Scott Reef and Rowley Reef Ecosystems in - Northwestern Australia
- **Temporal Coverage**: 1994 to 2020
- **Parameters of Interest for Analysis:** Species Pop

**2.** NOAA/ National Weather Service National Centers for Environmental Prediction Climate Prediction Center Oceanic Niño Index (ONI).<br> [NOAA ENSO Data](https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt)
- **Download Method:** read.table from base R
- **Spatial Coverage**: 5N to 5S, 170W to 120W
- **Temporal Coverage**: 1950 to 2021
- **Parameters of Interest for Analysis:** SST Anomaly 


Our group used a KNB project repository to preserve our data due to their reputation, reliability, and commitment to the future in case of company changes.
Data will be preserved until the Simpson’s Diversity Index methodology has been superseded by more significant research. 

**Link to KNB Repository:**
[KNB Ecoinformatic site](https://knb.ecoinformatics.org/)


